Renewal and stability in populations structured by remaining years of life

### 2013-11-15 ThanoRepro/ThanoRepro.tex
.

\begin{align}
\label{eq:transform}
P_y =& \sum_{x=0}^\omega P_x \frac{d_{x+y}}{l_x}
\intertext{and of course}
P =& \sum P_y = \sum P_a
\end{align}

We refer to \eqref{eq:transform} as the thanatological age
transformation\footnote{I have on good authority that Ken Wachter
coined the term thanatological age.}, and this complements the notion of
chronological age. It bears repeating that thanatological age is merely probabilistic, since it
refers to the future according to particular assumptions about the survival
curve. As such, Figure~\ref{fig:USrecomp} is projective in nature, a
forward-looking glance at a population given the age-structure and lifetable of a particular moment, whereas
Figure~\ref{fig:USdecomp} is reflective in nature, since a population's
age-structure is mostly the fruit of past fertility, but also migration and to a
lesser degree attrition. In a different paper we explore population structure by remaining years of life
for its own sake. Note that any age-classified count can be
reclassified in this way, as long as you have a lifetable that plausibly
represents the population whose data you wish to reclassify. This means we can
derive thanatological fertility rates, $F_y$, by applying \eqref{eq:transform}
to age-classified birth counts, $B_x$, to get $B_y$ and again to
exposure-to-risk, $E_x$ (here we take exposures from all ages), to get $E_y$ and then dividing:

@@

\section*{The thanatological renewal model}
For the population model that follows, we are most interested in using the kind
of fertility rates shown in Figure~\ref{fig:FySpaghetti}, and we refer to a
unisex population. Assuming constant vital rates, the births for the present year are given by

\begin{align}
B(t) =& \int _0^\infty P(y,t)F(y) \dd y = \int _0^\infty P(a,t)F(a) \dd a
\intertext{where $P(a), P(y)$ are population counts and $F(y), F(a)$ are exact
specific fertility probabilities (rates), and $d(a)$ is the continuous
lifetable death distribution with radix of 1, so $d(a)/l(0)$. The
thanatological integral can be broken down back in terms of chronological age.} =&
\int_{y=0}^\infty
\int_{a=0}^\infty F(y)\frac{P(a,t)d_(a+y)}{l(a)} \dd a \dd y 
\intertext{We can relate the present population to past
births with $P(a,t) = B(t-a)l(a)$.} 
\label{eq:ergostep1}
=& \int_{y=0}^\infty \int_{a=0}^\infty F(y) B(t-a)d(a+y)\dd a \dd y
\intertext{And eventually-- a bit quicker than is the case for chronological
age-- strong ergodicity will assert itself, and $B(t)$ will be related to
$B(t-a)$ according to a constant factor $e^{ra}$, where $r$ is the familiar
intrinsic rate of growth.}
\label{eq:ergostep2}
=& \int_{y=0}^\infty \int_{a=0}^\infty F(y) B(t)e^{-ra}d(a+y)\dd a \dd y
\intertext{Divide out $B(t)$ to get back a familiar-looking renewal equation.}
\label{eq:thanoren}
1 =& \int_{y=0}^\infty \int_{a=0}^\infty F(y) d(a+y)e^{-ra}\dd a \dd y
\end{align}
Now compare this to Lotka's chronological formulation
\begin{equation}
\label{eq:lotka}
1 = \int_{a=0}^\infty F(a)l(a)e^{-ra}\dd a
\end{equation}
and note that these are really quite similar, since $\int _{y=0}^\infty
d(a+y)\dd a = l(a)$. For intuition, notice that $l(a)$ is here split
up into pieces of $d(a)$, and imagine a 2D surface of these, where one axis is chronological age and the
other axis is thanatological age. For the chronological case \eqref{eq:lotka},
we multiply chronological age-specific fertility rates over one margin, and for
the thanatological case over the other margin. There is a strong parallel here
with the case of two-sex age-specific fertility rates. As in the case of
divergence between male and female single-sex models, there will always be divergence between
single-sex models under chronological versus thanatological age. Even though the
modeled population stocks are in a way commensurable, the rates used, $F(a)$
versus $F(y)$ are calculated on the basis of differently distributed
denominators $E(a)$ versus $E(y)$.

There are gaps in the above line of development, since the jump from
\eqref{eq:ergostep1} to \eqref{eq:ergostep2} (strong ergodicity) is unproven. A
further conjecture is that the jump will also hold for the case of weak
ergodicity. The mechanisms at play unfold in the same way as those so
intuitively described by \citet{arthur1982ergodic}, and said proof may apply
here without further modification. A proof of
the uniqueness of the solution to \eqref{eq:thanoren} is given in Appendix~\ref{app:A}.

@@

What should one imagine under the model of thanatological population renewal? A
useful mnemonic bases itself on Figures~\ref{fig:USdecomp} and
\ref{fig:USrecomp}. In the age-structured model, new generations
appear at the bottom of the pyramid, and move up one rung per year. All ages are subject to decrement. In the thanatological \textit{leaf}, each birth cohort increments to
the population over the whole range of thanatological age according to $d(a)$,
as seen in \eqref{eq:thanoren}. Birth cohorts become the layers seen in
Figure~\ref{fig:USrecomp}. Each horizontal step is a death cohort, and these
move one step down the pyramid each year without any decrement (indeed
incrementing) until reaching the very bottom. In short, the locations of
increment and decrement, and the direction of movement are all switched.

@@

@@

[\ldots paper to be updated and explanded\ldots]
\nocite{HMD,HFD}
% --------------------------------------------------
% bibliography
\bibliographystyle{plainnat}
  \bibliography{References}

>>>>

\begin{align}
\label{eq:transform}
P_y =& \sum_{x=0}^\omega P_x \frac{d_{x+y}+d_{x+y+1}}{l_x+l_{x+1}}
\intertext{and of course}
P =& \sum P_y = \sum P_a
\end{align}

We refer to \eqref{eq:transform} as the thanatological age
transformation\footnote{I have on good authority that Ken Wachter
coined the term \textit{thanatological} age.}, and this complements the notion
of chronological age. It bears repeating that thanatological age is merely probabilistic, since it
refers to the future according to particular assumptions about the survival
curve. As such, Figure~\ref{fig:USrecomp} is projective in nature, a
forward-looking glance at a population given the age-structure and lifetable of a particular moment, whereas
Figure~\ref{fig:USdecomp} is reflective in nature, since a population's
age-structure is mostly the fruit of past fertility, but also migration and to a
lesser degree attrition. In a different paper we explore population structure by remaining years of life
for its own sake. Note that any age-classified count can be
reclassified in this way, as long as you have a lifetable that plausibly
represents the population whose data you wish to reclassify. This means we can
derive thanatological fertility rates, $F_y$, by applying \eqref{eq:transform}
to age-classified birth counts, $B_x$, to get $B_y$ and again to
exposure-to-risk, $E_x$ (here we take exposures from all ages), to get $E_y$ and then dividing:

@@

In general, such fertility rates will not be palatable for purposes of
projection, unless the demographer believes that their empirical regularity is
somehow stronger than typical ASFR --- an argument we do not make. There may be
alternative ways of defining thanatological fertility rates, for instance
\textit{after} stability, which would make this model component more conformable
with the Lotka renewal model. First let us define the basic model.

\section*{The thanatological renewal model}
For the population model that follows, we are most interested in using the kind
of fertility rates shown in Figure~\ref{fig:FySpaghetti}, and we refer to a
unisex population. Assuming constant vital rates, the births for the present
year are given by:
\begin{align}
B(t) =& \int _0^\infty P(y,t)F(y) \dd y = \int _0^\infty P(a,t)F(a) \dd a
\intertext{where $P(a), P(y)$ are population counts and $F(y), F(a)$ are exact
specific fertility probabilities (rates), and $d(a)$ is the continuous
lifetable death distribution with radix of 1, so $d(a)/l(0)$. The
thanatological integral can be broken down back in terms of chronological age.} =&
\int_{y=0}^\infty
\int_{a=0}^\infty F(y)\frac{P(a,t)d_(a+y)}{l(a)} \dd a \dd y 
\intertext{We can relate the present population to past
births with $P(a,t) = B(t-a)l(a)$.} 
\label{eq:ergostep1}
=& \int_{y=0}^\infty \int_{a=0}^\infty F(y) B(t-a)d(a+y)\dd a \dd y
\intertext{And eventually-- a bit quicker than is the case for chronological
age-- strong ergodicity will assert itself, and $B(t)$ will be related to
$B(t-a)$ according to a constant factor $e^{ra}$, where $r$ is the familiar
intrinsic rate of growth.}
\label{eq:ergostep2}
=& \int_{y=0}^\infty \int_{a=0}^\infty F(y) B(t)e^{-ra}d(a+y)\dd a \dd y
\intertext{Divide out $B(t)$ to get back a familiar-looking renewal equation.}
\label{eq:thanoren}
1 =& \int_{y=0}^\infty \int_{a=0}^\infty F(y) d(a+y)e^{-ra}\dd a \dd y
\end{align}
Now compare this to Lotka's chronological formulation
\begin{equation}
\label{eq:lotka}
1 = \int_{a=0}^\infty F(a)l(a)e^{-ra}\dd a
\end{equation}
and note that these are really quite similar, since $\int _{y=0}^\infty
d(a+y)\dd a = l(a)$. For intuition, notice that $l(a)$ is here split
up into pieces of $d(a)$, and imagine a 2D surface of these, where one axis is chronological age and the
other axis is thanatological age. For the chronological case \eqref{eq:lotka},
we multiply chronological age-specific fertility rates over one margin, and for
the thanatological case over the other margin. There is a strong parallel here
with the case of two-sex age-specific fertility rates. As in the case of
divergence between male and female single-sex models, there will always be divergence between
single-sex models under chronological versus thanatological age. Even though the
modeled population stocks are in a way commensurable, the rates used, $F(a)$
versus $F(y)$ are calculated on the basis of differently distributed
denominators $E(a)$ versus $E(y)$.

There are gaps in the above line of development, since the jump from
\eqref{eq:ergostep1} to \eqref{eq:ergostep2} (strong ergodicity) is unproven,
although it is rather intuitive, given the much greater density of connections
within the model; persons from nearly any thanatologcal age can produce
offspring that can have any other thanatological age. In this sense, the
smoothing mechanism must be much stronger than that for chronoloical age, at
least in most cases. A further conjecture is that the jump will also hold for
the case of weak ergodicity. The mechanisms at play unfold in the same way as those so intuitively described by \citet{arthur1982ergodic}, and said proof may apply here without further modification. A proof of the uniqueness of the solution to \eqref{eq:thanoren} is given in Appendix~\ref{app:A}.

@@

The thanatological renewal model is coherent, but the idea may seem strange.
What should one imagine under the model of thanatological population renewal? 
A useful mnemonic bases itself on Figures~\ref{fig:USdecomp} and
\ref{fig:USrecomp}. In the age-structured model, new generations
appear at the bottom of the pyramid, and move up one rung per year. All ages 
are subject to decrement. In the thanatological \textit{leaf}, each birth cohort increments to
the population over the whole range of thanatological age according to $d(a)$,
as seen in \eqref{eq:thanoren}. Birth cohorts become the layers seen in
Figure~\ref{fig:USrecomp}. Each horizontal step is a death cohort, and these
move one step down the pyramid each year without any decrement (indeed
incrementing) until reaching the very bottom. In short, the locations of
increment and decrement, and the direction of movement are all switched.

@@

* this is a work in progress, and by the time the EPC comes around I'll
have shifted the focus away on comparing stable results estimated from data
(the last couple figures), and more toward the reversibility of chronological
and thanatological stable age structures, as well as a comparative
eigen-analysis of the Leslie and thanatological projection matrices.

@@

\nocite{HMD,HFD}
% --------------------------------------------------
% bibliography
\bibliographystyle{plainnat}
  \bibliography{References}
